---
title: "*Spatial Statistics Lab 6*"
author: '**Onyedikachi J Okeke**'
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### **0.0 To Load the library**
```{r,comment="",echo=TRUE,eval=FALSE}
library(sp)
library(spatstat)
library(sf)
library(spatstat.geom)
library(ctv)
library(terra)
library(spdep)
library(rgdal)
library(rgdal)
library(terra)
library(RColorBrewer)
library(classInt)

```
##### Library
```{r,echo=F,results='Show', fig.show='hold',eval=T}
library(sp)
library(spatstat)
library(sf)
library(spatstat.geom)
library(ctv)
library(terra)
library(spdep)
library(rgdal)
library(rgdal)
library(terra)
library(RColorBrewer)
library(classInt)

```

### **Q1 Loading Sids data**
```{r,comment="",echo=TRUE,eval=FALSE}
# Define the file path
file_path <- "C:/Spatial Statistics Labwork/Lab5Data/sids2.shp"

# Import the shapefile
sids <- st_read(file_path)

class(sids)

# Check the structure of the imported object
str(sids)

```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}
# Define the file path
file_path <- "C:/Spatial Statistics Labwork/Lab5Data/sids2.shp"

# Import the shapefile
sids <- st_read(file_path)

class(sids)

# Check the structure of the imported object
str(sids)

```

### **Check the projection of the shapefile**
```{r,comment="",echo=TRUE,eval=FALSE}
# Extract CRS information using st_crs()
crs_info <- st_crs(sids)

# Print CRS information
print(crs_info)

```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}
# Extract CRS information using st_crs()
crs_info <- st_crs(sids)

# Print CRS information
print(crs_info)

```

### **Define the target CRS (NAD27 UTM Zone 17N**
```{r,comment="",echo=TRUE,eval=FALSE}

# Assign the coordinate system (WGS84)
st_crs(sids) <- st_crs("+proj=longlat +ellps=WGS84")

# Check the current CRS
print(st_crs(sids))

# Define the target CRS (NAD27 UTM Zone 17N)
target_crs <- "+proj=utm +zone=17 +datum=NAD27"

# Project the shapefile to the target CRS
sids_projected <- st_transform(sids, target_crs)

# Check the CRS of the projected shapefile
print(st_crs(sids_projected))

```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}

# Assign the coordinate system (WGS84)
st_crs(sids) <- st_crs("+proj=longlat +ellps=WGS84")

# Check the current CRS
print(st_crs(sids))

# Define the target CRS (NAD27 UTM Zone 17N)
target_crs <- "+proj=utm +zone=17 +datum=NAD27"

# Project the shapefile to the target CRS
sids_NAD <- st_transform(sids, target_crs)

# Check the CRS of the projected shapefile
print(st_crs(sids_NAD))

```

### **#difine contiguity based neighbors**
```{r,comment="",echo=TRUE,eval=FALSE}
# Create Queen contiguity spatial weights matrix
sids_nbq <- poly2nb(sids, queen = TRUE)
# Check the summary of the spatial weights matrix
summary(sids_nbq)

# Create Queen (false) contiguity spatial weights matrix
sids_nbr <- poly2nb(sids, queen = FALSE)
# Check the summary of the spatial weights matrix
summary(sids_nbr)

coords <- st_coordinates(sids)

# Check the number of neighborhoods in sids_nbq
num_neighborhoods <- length(sids_nbq)

# Check the number of coordinate pairs in coords
num_coords <- nrow(coords)

# Print out the counts for verification
print(paste("Number of neighborhoods:", num_neighborhoods))
print(paste("Number of coordinate pairs:", num_coords))
```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}
# Create Queen contiguity spatial weights matrix
sids_nbq <- poly2nb(sids, queen = TRUE)
# Check the summary of the spatial weights matrix
summary(sids_nbq)

# Create Queen (false) contiguity spatial weights matrix
sids_nbr <- poly2nb(sids, queen = FALSE)
# Check the summary of the spatial weights matrix
summary(sids_nbr)

coords <- st_coordinates(sids)

# Check the number of neighborhoods in sids_nbq
num_neighborhoods <- length(sids_nbq)

# Check the number of coordinate pairs in coords
num_coords <- nrow(coords)

# Print out the counts for verification
print(paste("Number of neighborhoods:", num_neighborhoods))
print(paste("Number of coordinate pairs:", num_coords))

```

### **Row-standardized weights matrix**

```{r,comment="",echo=TRUE,eval=FALSE}
#Row-standardized weights matrix
sids_nbq_w<- nb2listw(sids_nbq)
print(sids_nbq_w)
sids_nbq_w$neighbours[1:5]

```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}
#Row-standardized weights matrix
sids_nbq_w<- nb2listw(sids_nbq)
print(sids_nbq_w)
sids_nbq_w$neighbours[1:5]

```

### **Binary Weights**
```{r,comment="",echo=TRUE,eval=FALSE}
#Binary Weights
sids_nbq_wb<-nb2listw(sids_nbq, style="B")
#examine the dataset
summary (sids)

```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}
#Binary Weights
sids_nbq_wb<-nb2listw(sids_nbq, style="B")
#examine the dataset
summary (sids)

```


### **Plot the SIDR74 from 1974-1978**
```{r,comment="",echo=TRUE,eval=FALSE}
# If `sids` is an sf object, convert it to SpatialPolygonsDataFrame or similar
if (class(sids)[1] == "sf") {
    library(sf)
    sids <- as(sids, "Spatial") # Converts sf object to sp class
}

if (!"SIDR74" %in% names(sids)) {
    stop("Column 'SIDR74' does not exist in 'sids'")
}

spplot(sids, "SIDR74")

```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}
# If `sids` is an sf object, convert it to SpatialPolygonsDataFrame or similar
if (class(sids)[1] == "sf") {
    library(sf)
    sids <- as(sids, "Spatial") # Converts sf object to sp class
}

if (!"SIDR74" %in% names(sids)) {
    stop("Column 'SIDR74' does not exist in 'sids'")
}

spplot(sids, "SIDR74")

```

### **Define a new lagged variable using row standardized weights**
```{r,comment="",echo=TRUE,eval=FALSE}

sids$SIDR74_lag1 <- lag(sids_nbq_w, sids$SIDR74)
lm2 <- lm(sids$SIDR74_lag1~sids$SIDR74)
lm2
plot(y= sids$SIDR74_lag1, x= sids$SIDR74)
title("The relationship between a variable \n and the lag of itself is moran's I")

```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}
sids$SIDR74_lag1 <- lag(sids_nbq_w, sids$SIDR74)
lm2 <- lm(sids$SIDR74_lag1~sids$SIDR74)
lm2
plot(y= sids$SIDR74_lag1, x= sids$SIDR74)
title("The relationship between a variable \n and the lag of itself is moran's I")

```

Q1. What is the coefficient for variable SIDR74. What does this coefficient mean?

### **Define a new lagged variable using row standardized weights**
```{r,comment="",echo=TRUE,eval=FALSE}

sids$SIDR74_lag1 <- lag(sids_nbq_w, sids$SIDR74)
lm2 <- lm(sids$SIDR74_lag1~sids$SIDR74)
lm2
plot(y= sids$SIDR74_lag1, x= sids$SIDR74)
title("The relationship between a variable \n and the lag of itself is moran's I")

```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}
sids$SIDR74_lag1 <- lag(sids_nbq_w, sids$SIDR74)
lm2 <- lm(sids$SIDR74_lag1~sids$SIDR74)
lm2
plot(y= sids$SIDR74_lag1, x= sids$SIDR74)
title("The relationship between a variable \n and the lag of itself is moran's I")

```

### **Global moran’s I test**
```{r,comment="",echo=TRUE,eval=FALSE}

moran.test (sids$SIDR74, sids_nbq_w, randomisation = FALSE) # Moran I test under normality
moran.test(sids$SIDR74, sids_nbq_wb) # Moran I test under randomization using binary weight

```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}

moran.test (sids$SIDR74, sids_nbq_w, randomisation = FALSE) # Moran I test under normality
moran.test(sids$SIDR74, sids_nbq_wb) # Moran I test under randomization using binary weight

```

Q2. What are the results (pattern)? Compare their test statistic, standard deviate, p-value, and variance.

### **Monte-Carlo simulation of Moran’s I**
```{r,comment="",echo=TRUE,eval=FALSE}

set.seed(1234)
permu<-moran.mc(sids$SIDR74, sids_nbq_w, nsim = 999)
permu
graph999 <- hist(permu$res,freq=TRUE,col="light blue",main="Permutation Test for Moran's I - 999 permutations")
lines(permu$statistic,max(graph999$counts),type="h",col="red",lwd=2)

```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}

set.seed(1234)
permu<-moran.mc(sids$SIDR74, sids_nbq_w, nsim = 999)
permu
graph999 <- hist(permu$res,freq=TRUE,col="light blue",main="Permutation Test for Moran's I - 999 permutations")
lines(permu$statistic,max(graph999$counts),type="h",col="red",lwd=2)

```

Q3.Explain the histogram.

### **local Empirical Bayes estimates**
```{r,comment="",echo=TRUE,eval=FALSE}

set.seed(1234)
EBImoran.mc(n = sids$SID74, x = sids$BIR74, sids_nbq_wb, nsim = 999)

```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}

set.seed(1234)
EBImoran.mc(n = sids$SID74, x = sids$BIR74, sids_nbq_wb, nsim = 999)

```

Q4. How is the result from this method different from previous results?

### **Explore the Correlogram**
```{r,comment="",echo=TRUE,eval=FALSE}

cor8<-sp.correlogram(sids_nbq, sids$SIDR74, order=8, method="I")
cor8
plot(cor8)

```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}

cor8<-sp.correlogram(sids_nbq, sids$SIDR74, order=8, method="I")
cor8
plot(cor8)

```

Q5.Explain the plot

### **Explore the local Moran’ I**
```{r,comment="",echo=TRUE,eval=FALSE}

#First plot the moran scatterplot
moran.plot(sids$SIDR74, sids_nbq_w)
```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}

#First plot the moran scatterplot
moran.plot(sids$SIDR74, sids_nbq_w)

```

Q6.Explain the plot

### **LISA**
```{r,comment="",echo=TRUE,eval=FALSE}

locm <- localmoran(sids$SIDR74, sids_nbq_w) #Calculate the local Moran’s I
locm
summary(locm)
sids$sSIDR74<-scale(sids$SIDR74) #Standardize it
sids$lag_sSIDR74<-lag(sids_nbq_w, sids$sSIDR74) #Define a lag variable
sids$sSIDR74
sids$lag_sSIDR74
sids$quad_sig <- NA #Define a new variable quad_sig and set the initial value to be NA
#Next define five types of results: High-High, Low-Low, High-Low, Low-High, and not significant
sids@data[(sids$sSIDR74>= 0 & sids$lag_sSIDR74>= 0) & (locm[,5]<= 0.05),"quad_sig"] <- 1
sids@data[(sids$sSIDR74<= 0 & sids$lag_sSIDR74<= 0) & (locm[,5]<= 0.05),"quad_sig"] <- 2
sids@data[(sids$sSIDR74>= 0 & sids$lag_sSIDR74<= 0) & (locm[,5]<= 0.05),"quad_sig"] <- 3
sids@data[(sids$sSIDR74<= 0 & sids$lag_sSIDR74>= 0) & (locm[,5]<= 0.05),"quad_sig"] <- 4
sids@data[locm[,5]> 0.05, "quad_sig"] <- 5

```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}

locm <- localmoran(sids$SIDR74, sids_nbq_w) #Calculate the local Moran’s I
locm
summary(locm)
sids$sSIDR74<-scale(sids$SIDR74) #Standardize it
sids$lag_sSIDR74<-lag(sids_nbq_w, sids$sSIDR74) #Define a lag variable
sids$sSIDR74
sids$lag_sSIDR74
sids$quad_sig <- NA #Define a new variable quad_sig and set the initial value to be NA
#Next define five types of results: High-High, Low-Low, High-Low, Low-High, and not significant
sids@data[(sids$sSIDR74>= 0 & sids$lag_sSIDR74>= 0) & (locm[,5]<= 0.05),"quad_sig"] <- 1
sids@data[(sids$sSIDR74<= 0 & sids$lag_sSIDR74<= 0) & (locm[,5]<= 0.05),"quad_sig"] <- 2
sids@data[(sids$sSIDR74>= 0 & sids$lag_sSIDR74<= 0) & (locm[,5]<= 0.05),"quad_sig"] <- 3
sids@data[(sids$sSIDR74<= 0 & sids$lag_sSIDR74>= 0) & (locm[,5]<= 0.05),"quad_sig"] <- 4
sids@data[locm[,5]> 0.05, "quad_sig"] <- 5

```

### **Plot the LISA Map**
```{r,comment="",echo=TRUE,eval=FALSE}

breaks <- seq(1, 5, 1)
labels <- c("High-High", "Low-Low", "High-Low", "Low-High", "Not Signif.")
np <- findInterval(sids$quad_sig, breaks)
colors <- c("red", "blue", "lightpink", "skyblue2", "white")
plot(sids, col = colors[np]) #colors[np] manually sets the color for each county
mtext("Local Moran's I", cex = 1.5, side = 3, line = 1)
legend("topleft", legend = labels, fill = colors, bty = "n")

```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}

breaks <- seq(1, 5, 1)
labels <- c("High-High", "Low-Low", "High-Low", "Low-High", "Not Signif.")
np <- findInterval(sids$quad_sig, breaks)
colors <- c("red", "blue", "lightpink", "skyblue2", "white")
plot(sids, col = colors[np]) #colors[np] manually sets the color for each county
mtext("Local Moran's I", cex = 1.5, side = 3, line = 1)
legend("topleft", legend = labels, fill = colors, bty = "n")

```

Q7. Plot and explain the pattern.

### **explore the Getis G**
```{r,comment="",echo=TRUE,eval=FALSE}
globalG.test(sids$SIDR74, sids_nbq_wb)

```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}
globalG.test(sids$SIDR74, sids_nbq_wb)
```

Q8. What are your results and explain the pattern (clustered? Hotspot or cold spot?)

### **Local Getis G**
```{r,comment="",echo=TRUE,eval=FALSE}
G <- localG(sids$SIDR74, sids_nbq_wb)
par.G <- par(mfrow=c(1,1))
# set the graphics parameters to display to maps on top of each other
# in the same window
brks <- seq(-4,6,1)
# creates 10 intervals between -4 and +5
# summary(G) the variable to be plotted has a low of -3.6 and high of 5.7
cm.col <- cm.colors(length(brks)-1)
# picks colors corresponding for each interval out of the "cm" color palette
plot(sids,border="lightgray",col=cm.col[findInterval(G,brks,all.inside=TRUE)])
# maps the shape file and fills each polygon with a color corresponding to
# the level of G at that location
text(coordinates(sids)[,1],coordinates(sids)[,2], round(G, digits=1), cex=0.5)
# writes the G-statistic in each polygon
title("Getis-Ord G statistic for sids")

```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}
G <- localG(sids$SIDR74, sids_nbq_wb)
par.G <- par(mfrow=c(1,1))
# set the graphics parameters to display to maps on top of each other
# in the same window
brks <- seq(-4,6,1)
# creates 10 intervals between -4 and +5
# summary(G) the variable to be plotted has a low of -3.6 and high of 5.7
cm.col <- cm.colors(length(brks)-1)
# picks colors corresponding for each interval out of the "cm" color palette
plot(sids,border="lightgray",col=cm.col[findInterval(G,brks,all.inside=TRUE)])
# maps the shape file and fills each polygon with a color corresponding to
# the level of G at that location
text(coordinates(sids)[,1],coordinates(sids)[,2], round(G, digits=1), cex=0.5)
# writes the G-statistic in each polygon
title("Getis-Ord G statistic for sids")
```

Q9.Is it different from the LISA map in Q7? Why?

### **Geary’s C**
```{r,comment="",echo=TRUE,eval=FALSE}
library(spdep)
geary.test(sids$SIDR74, sids_nbq_wb, randomisation=FALSE)
gc_perm <- geary.mc(sids$SIDR74, sids_nbq_wb, nsim=99)
gc_perm

```
##### Result
```{r echo=FALSE, results='show', fig.show='hold', eval=TRUE, warning=FALSE, fig.width=6, fig.height=4}
library(spdep)
geary.test(sids$SIDR74, sids_nbq_wb, randomisation=FALSE)
gc_perm <- geary.mc(sids$SIDR74, sids_nbq_wb, nsim=99)
gc_perm

```

Q10. What is the pattern? Are these results different? Explain the differences if any.


